\chapter*{\textbf{ABSTRACT}}
\addcontentsline{toc}{chapter}{ABSTRACT}

\noindent
Privacy regulations such as the General Data Protection Regulation (GDPR) and the California Consumer Privacy Act (CCPA) grant users the legal right to request deletion of their personal data from machine learning systems. Honoring such requests is technically non-trivial, as trained models implicitly encode the influence of every data point seen during training. Retraining from scratch on every deletion request is computationally prohibitive, particularly as dataset sizes and model complexity grow.

\noindent
This project presents PyUnlearn, a practical machine unlearning SDK and API platform built on the Sharded, Isolated, Sliced, and Aggregated (SISA) framework. By partitioning training data into independently retrainable shards, the system localizes the impact of any deletion request to a single shard, avoiding full model retraining. The system is designed to be compatible with any scikit-learn estimator, including Linear Regression, Logistic Regression, Support Vector Machines, Random Forests, and Gradient Boosting models, making it broadly applicable across classical machine learning workflows.

\noindent
The architecture incorporates persistent SQLite-based metadata storage for stable index-to-shard mappings, an append-only audit log for regulatory traceability, and idempotent unlearning semantics to ensure operational safety under repeated deletion requests. An optional non-blocking background synchronization layer enables real-time monitoring through a Django backend and React-based dashboard without disrupting core training or unlearning operations. The system is packaged as a reusable SDK with configuration-driven deployment supporting both local and cloud-enabled environments.

\noindent
Performance benchmarks on datasets ranging from 5,000 to 50,000 samples demonstrate significant speedup factors over full retraining, with accuracy retention maintained across varying shard configurations. Remaining work includes extended support for deep learning architectures, additional dataset validation, and completion of the production deployment pipeline.

\vspace{0.8cm}
\noindent\textit{\textbf{Keywords:} Machine Unlearning, SISA Framework, Right to be Forgotten, GDPR, Privacy-Compliant Machine Learning, Selective Data Deletion, Federated Unlearning, scikit-learn, SDK, Audit Logging, Sharded Training, Data Privacy}
