\chapter*{\textbf{ACKNOWLEDGEMENT}}
\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENT}

We wish to express our heartfelt gratitude to everyone who has contributed to the successful realization of this project.

We are immensely grateful to the Department of Electronics and Computer Engineering, Purwanchal Campus, Institute of Engineering, Tribhuvan University, and the Head of Department, Asst. Prof. Manoj Kumar Guragain, for providing us with the essential resources, infrastructure, and opportunities to bring our ideas to life.

We would like to express our sincere gratitude to our respected project supervisor, Asst. Prof. Pukar Karki, for his invaluable guidance, encouragement, and continuous support throughout the course of this project. His expert advice, constructive feedback, and keen insights have been instrumental in shaping our work and ensuring its successful completion.

We also extend our appreciation to all the teachers and staff of the department for their guidance, mentorship, and assistance, which have been instrumental in supporting our academic and project journey. We are equally thankful to our colleagues and peers for their constructive feedback, collaborative brainstorming sessions, and moral support, which kept us motivated and inspired during the more challenging moments of this journey.

Lastly, we would like to express our deepest gratitude to our families and friends for their steadfast support, patience, and encouragement, which were invaluable throughout this endeavor.

This project stands as a testament to the collective efforts and contributions of everyone mentioned above. We are truly thankful for their support and inspiration.

\vspace{0.5cm}
\noindent Thank You!

