\pagenumbering{arabic}
\chapter{INTRODUCTION}

\section{Background}
Machine learning models have become deeply integrated into everyday applications, from personalized content recommendations to critical tasks like medical diagnosis and financial fraud detection. These models learn patterns from vast amounts of user data, improving their accuracy and usefulness over time. However, this dependence on user data creates a fundamental tension with privacy rights. Regulations such as the General Data Protection Regulation (GDPR) in Europe and the California Consumer Privacy Act (CCPA) in the United States now establish clear legal requirements: users have the right to request deletion of their personal data from systems that process it. Beyond these two landmark regulations, similar legislation such as PIPEDA in Canada and the Personal Data Protection Act (PDPA) in various Asian jurisdictions further reinforce the global momentum toward data sovereignty and the right to be forgotten.

The technical challenge lies in how to honor such deletion requests when the data has already been incorporated into a trained machine learning model. Because ML models potentially memorize training data during the learning process, simply removing a record from a database does not eliminate its influence from a deployed model. This memorization effect is particularly pronounced in overparameterized models such as deep neural networks, which have sufficient capacity to encode individual training examples. As a result, a model trained on sensitive data may remain vulnerable to membership inference attacks and other privacy exploits even after the underlying data has been deleted from storage.

Traditional approaches require complete model retraining on the remaining dataset, a process that becomes prohibitively expensive as model complexity and dataset sizes grow. Retraining a large model could take hours or days, making it impossible to honor deletion requests within reasonable timeframes, especially when multiple deletion requests arrive frequently. Furthermore, in online learning settings where models are continuously updated with new data, the cost of repeated full retraining compounds significantly over time.

This gap between legal requirements and technical capability has given rise to the field of machine unlearning, which aims to efficiently remove the influence of specific data points from trained models without full retraining. Unlearning guarantees that training on a point and subsequently unlearning it will produce the same distribution of models as not training on that point at all. This strong guarantee is what distinguishes machine unlearning from related privacy-preserving techniques such as differential privacy, which only bounds the contribution of individual points rather than eliminating it entirely.

\section{Gap Identification}
Despite growing research interest in machine unlearning, the practical adoption of these techniques remains limited. Academic papers propose sophisticated algorithms, but they often remain as theoretical prototypes or are tightly coupled to specific model architectures. Prior work has shown that exact unlearning is possible when the learning algorithm queries data in a fixed, predetermined order, as in the statistical query (SQ) learning setting. However, when the dataset is queried adaptively, convergence of such approaches is no longer guaranteed, and efficient algorithms for complex models such as deep neural networks do not exist in this setting.

Approximate unlearning methods, including influence functions and knowledge distillation on cleaned datasets, offer alternatives but introduce their own limitations. Influence functions require access to second-order gradient information, which is computationally expensive for large models. Knowledge distillation approaches depend on the availability of a suitable proxy dataset and do not provide formal guarantees of data removal. These constraints make them difficult to deploy reliably in production environments where regulatory compliance demands verifiable and auditable deletion.

Open-source libraries that do exist tend to focus on narrow use cases or require significant expertise to integrate into existing machine learning pipelines. For most developers and organizations, there is no straightforward, production-ready solution that allows them to build privacy-compliant systems with unlearning capabilities from day one. This project addresses that gap by developing PyUnlearn, a practical SDK and API platform that abstracts away the complexity of machine unlearning, making it accessible to developers working on real-world applications that must comply with privacy regulations.

\section{Motivation}
The motivation for this project stems from a simple but increasingly urgent question: as users entrust their data to machine learning systems, can they also trust that their presence can be removed when requested? Current practices often treat data deletion as a database operation, ignoring the fact that the model itself retains implicit knowledge of the deleted data. This disconnect between what users expect and what systems actually deliver creates legal risk for organizations and erodes user trust.

The SISA (Sharded, Isolated, Sliced, and Aggregated) training framework, introduced by Bourtoule et al., provides a structured and practical approach to this problem. By dividing training data into multiple disjoint shards and training models in isolation on each shard, the influence of any single data point is limited to the model trained on the shard containing that point. When an unlearning request arrives, only the affected shard needs to be retrained, dramatically reducing computational overhead compared to full retraining and making it feasible to honor deletion requests within practical timeframes.

Rather than treating privacy as an afterthought or a compliance checkbox, we believe that unlearning capabilities should be built directly into the machine learning development workflow. PyUnlearn represents an effort to make this possible, providing developers with tools that integrate naturally with existing scikit-learn-based workflows while ensuring that privacy guarantees are maintained throughout the model lifecycle. The system further extends the SISA framework with production-grade features including persistent metadata storage, audit logging, idempotent deletion semantics, and an optional real-time monitoring dashboard.

\section{Objectives}
The project aims to achieve the following objectives:
\begin{itemize}
\item Develop an efficient SISA-based selective unlearning SDK compatible with any scikit-learn estimator, enabling localized retraining without full model rebuilding.
\item Ensure regulatory compliance by providing persistent audit logging, idempotent deletion semantics, and verifiable unlearning behavior aligned with GDPR and CCPA requirements.
\item Deliver a production-ready platform with a Django and React monitoring dashboard, supporting both local and cloud-enabled deployments through configuration-driven setup.
\end{itemize}

\section{Scope and Limitations}
This project focuses on implementing a practical machine unlearning SDK based on the SISA framework, targeting tabular datasets and machine learning models compatible with the scikit-learn ecosystem. The system supports both local deployment and cloud-enabled monitoring through an optional dashboard built with Django and React. The SDK is designed to handle both individual and batch deletion requests, with persistent SQLite-based metadata ensuring stable shard mappings and full auditability across repeated executions. Configuration-driven deployment allows users to switch between local-only and cloud-synchronized modes without modifying their core machine learning code, making the system adaptable to a wide range of development and production environments.

While the approach demonstrates significant efficiency gains over full retraining, current limitations include the lack of support for deep learning architectures such as convolutional neural networks. The SISA framework achieves the largest improvements for stateful algorithms like stochastic gradient descent used in deep neural networks; however, adapting it to such architectures introduces additional complexity around gradient propagation and initialization strategies that are beyond the current scope. The system also relies on a fixed sharding strategy that may not adapt optimally to all dataset distributions, and the optimal shard count must currently be determined empirically rather than derived automatically from dataset characteristics. Furthermore, formal certification guarantees for the unlearning process are not provided, meaning the system offers practical rather than mathematically provable privacy assurances. These limitations represent well-defined directions for future research and engineering work beyond the scope of the current project.
