\chapter{LITERATURE REVIEW}

Machine learning models have become central to a wide range of applications, from personalized recommendation systems to healthcare diagnostics and financial fraud detection. These models are trained on large volumes of user-generated data, and their predictive power is directly tied to the richness of that data. However, this reliance on personal data raises significant privacy concerns. Regulations such as the General Data Protection Regulation (GDPR) \cite{gdpr2016} and the California Consumer Privacy Act (CCPA) legally mandate that individuals have the right to request deletion of their personal data from any system that processes it, including trained machine learning models. The scale at which machine learning is now being applied to personal data makes the technical implementation of this right both urgent and non-trivial.

The concept of machine unlearning was formally introduced by Cao and Yang \cite{cao2015towards}, who defined it as the process of making a system forget a subset of its training data. Their work established the foundational requirement that an unlearned model must be statistically indistinguishable from a model that was never trained on the forgotten data. This definition set the standard for subsequent research and highlighted the inadequacy of simply deleting records from a database without addressing the model's retained knowledge. The naive solution to unlearning is complete model retraining on the remaining dataset after each deletion request. While this guarantees exact removal, it is computationally prohibitive for large-scale systems. As noted by Ginart et al. \cite{ginart2019making}, the cost of retraining grows with dataset size and model complexity, making it infeasible for production systems that receive frequent deletion requests. Their work also introduced the concept of data deletion in the context of $k$-means clustering, demonstrating that efficient unlearning is achievable for certain model families without full retraining.

A key motivation for machine unlearning is the vulnerability of trained models to privacy attacks. Shokri et al. \cite{shokri2017membership} demonstrated that machine learning models are susceptible to membership inference attacks, in which an adversary can determine with high confidence whether a specific data point was part of the training set. Their work showed that models tend to produce higher confidence predictions on training data than on unseen data, a property that can be exploited to infer membership. This overfitting-based signal is particularly strong in deep neural networks trained without sufficient regularization. Beyond membership inference, model inversion attacks can reconstruct approximate representations of training data from model outputs, and data extraction attacks can recover verbatim training examples from generative models. These attack vectors collectively demonstrate that a deployed model retains implicit knowledge of its training data long after the source data has been deleted from storage, making effective unlearning mechanisms essential for genuine privacy compliance.

The most prominent exact unlearning framework is SISA training, proposed by Bourtoule et al. \cite{bourtoule2021machine}. SISA partitions the training dataset into multiple disjoint shards and trains an independent sub-model on each shard. When a deletion request arrives, only the shard containing the target data point needs to be retrained, leaving all other shards untouched. The framework further introduces slicing, where each shard is divided into time-ordered slices and model checkpoints are saved after each slice. When unlearning is required, retraining begins from the last clean checkpoint rather than from random initialization, providing an additional reduction in retraining time. At inference time, predictions from all shard models are aggregated through majority voting for classification or averaging for regression. Bourtoule et al. established SISA as the state-of-the-art approach for exact unlearning in practical settings, demonstrating its effectiveness across multiple datasets and model architectures \cite{bourtoule2021machine}.

Approximate unlearning methods trade off exactness for reduced computational cost. Guo et al. \cite{guo2020certified} proposed certified data removal, which uses Newton update steps to approximately remove the influence of a data point from trained model parameters, providing $(\epsilon, \delta)$-certified removal guarantees. While effective for convex models such as logistic regression, the approach requires second-order gradient information, limiting its scalability to large neural networks. Golatkar et al. \cite{golatkar2020eternal} addressed selective forgetting in deep neural networks through a scrubbing procedure based on the Fisher information matrix, adding structured noise to model weights to mask the influence of forgotten data. Their approach is particularly relevant for deep learning architectures where exact retraining is computationally expensive, and it provides a principled way to balance forgetting quality against model utility on retained data. Warnecke et al. \cite{warnecke2021machine} further investigated unlearning at the level of individual features and labels rather than entire data points, demonstrating that feature-level unlearning can be more efficient when forgotten data shares features with retained data. This granularity of unlearning is relevant for scenarios where users request removal of specific attributes rather than entire records.

Sekhari et al. \cite{sekhari2021remember} extended the theoretical analysis of unlearning by examining its connection to the generalization properties of learning algorithms, showing that efficient unlearning is possible when the learning algorithm satisfies certain stability conditions. Their work provided sample complexity bounds for unlearning, establishing that the number of samples required to achieve a given unlearning quality scales with the complexity of the model class. In specialized domains, Chen et al. \cite{chen2022recommendation} investigated unlearning in recommendation systems, where user interaction histories are encoded in shared item embeddings that are jointly influenced by many users. They proposed a recalibration approach that selectively updates affected embeddings without full model retraining, addressing the unique challenge that a single user's data is distributed across many model parameters in collaborative filtering architectures.

Beyond algorithmic efficiency, Schelter et al. \cite{schelter2021heterogeneous} highlighted that production machine learning systems require persistent metadata management, versioned model artifacts, and auditable deletion records for reproducible and compliant unlearning operations. Their survey identified the lack of standardized tooling for data lifecycle management as a critical barrier to practical unlearning deployment. A production-ready unlearning system must maintain stable mappings between data indices and model partitions across system restarts, provide an append-only audit log of all deletion actions for regulatory compliance, and handle repeated unlearning requests idempotently without corrupting model state. These operational requirements are frequently overlooked in academic unlearning research, which tends to focus on algorithmic efficiency and theoretical guarantees while treating system-level concerns as implementation details. The UCI Machine Learning Repository \cite{dua2019uci} provides standardized benchmark datasets, including the Adult Income dataset used in our performance evaluation, enabling reproducible comparison of unlearning methods across different research groups and implementation approaches.

A review of existing systems reveals several recurrent limitations that this project directly addresses. First, most unlearning implementations are tightly coupled to specific model architectures or training frameworks, making them difficult to integrate into existing machine learning pipelines that use standard libraries such as scikit-learn. Second, existing tools rarely provide persistent state management, meaning that unlearning history and shard assignments are lost between system restarts, preventing reliable auditing and reproducibility. Third, the absence of developer-friendly APIs and configuration-driven deployment options creates a high barrier to adoption for organizations that lack deep expertise in privacy-preserving machine learning. Fourth, monitoring and observability features are largely absent from existing unlearning tools, making it difficult for operators to track unlearning events, verify deletion outcomes, and demonstrate compliance to regulators in real time.

Existing approaches each offer distinct advantages: SISA \cite{bourtoule2021machine} provides computational efficiency through localized retraining, influence functions \cite{guo2020certified} enable parameter-level updates without retraining, and certified removal offers formal privacy guarantees. However, no single solution comprehensively addresses the full set of operational requirements for real-world deployment, including broad model compatibility, persistent state management, regulatory audit logging, and developer-friendly integration. PyUnlearn bridges this gap by implementing the SISA framework with production-grade system features including SQLite-based metadata storage, append-only unlearning history logging, idempotent deletion semantics, and an optional real-time monitoring dashboard built with Django and React. By packaging these capabilities as a reusable SDK compatible with any scikit-learn estimator, PyUnlearn makes privacy-compliant machine learning accessible to developers without requiring deep expertise in unlearning algorithms, directly addressing the recurrent limitations identified across the existing literature.
