\begin{thebibliography}{1}

\bibitem{bourtoule2021machine}
L.~Bourtoule, V.~Chandrasekaran, C.~A.~Choquette-Choo, H.~Jia, A.~Travers, B.~Zhang, D.~Lie, and N.~Papernot, ``Machine unlearning,'' in \emph{IEEE Symposium on Security and Privacy (SP)}, 2021.

\bibitem{cao2015towards}
Y.~Cao and J.~Yang, ``Towards making systems forget with machine unlearning,'' in \emph{IEEE Symposium on Security and Privacy}, 2015.

\bibitem{ginart2019making}
A.~Ginart, M.~Guan, G.~Valiant, and J.~Y.~Zou, ``Making AI forget you: Data deletion in machine learning,'' in \emph{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 32, 2019.

\bibitem{guo2020certified}
C.~Guo, T.~Goldstein, A.~Hannun, and L.~van der Maaten, ``Certified data removal from machine learning models,'' in \emph{International Conference on Machine Learning (ICML)}, vol. 119, 2020.

\bibitem{chen2022recommendation}
C.~Chen, F.~Sun, M.~Zhang, and B.~Ding, ``Recommendation unlearning,'' in \emph{Proceedings of the Web Conference (WWW)}, 2022.

\bibitem{sekhari2021remember}
A.~Sekhari, J.~Acharya, G.~Kamath, and A.~T.~Suresh, ``Remember What You Want to Forget: Algorithms for Machine Unlearning,'' \emph{arXiv preprint arXiv:2103.03279}, 2021.

\bibitem{golatkar2020eternal}
A.~Golatkar, A.~Achille, and S.~Soatto, ``Eternal Sunshine of the Spotless Net: Selective Forgetting in Deep Networks,'' in \emph{Proceedings of the IEEE/CVF Conference on Computer Vision and Pattern Recognition (CVPR)}, 2020.

\bibitem{shokri2017membership}
R.~Shokri, M.~Stronati, C.~Song, and V.~Shmatikov, ``Membership Inference Attacks against Machine Learning Models,'' in \emph{IEEE Symposium on Security and Privacy (SP)}, 2017.

\bibitem{gdpr2016}
European Parliament and Council of the European Union, ``Regulation (EU) 2016/679 (General Data Protection Regulation),'' \emph{Official Journal of the European Union}, L 119, pp. 1--88, 4 May 2016.

\bibitem{dua2019uci}
D.~Dua and C.~Graff, ``UCI Machine Learning Repository,'' University of California, Irvine, School of Information and Computer Sciences, 2019.

\bibitem{warnecke2021machine}
A.~Warnecke, L.~Pirch, C.~Wressnegger, and K.~Rieck, ``Machine unlearning of features and labels,'' in \emph{Proceedings of the Network and Distributed System Security Symposium (NDSS)}, 2021.

\bibitem{schelter2021heterogeneous}
S.~Schelter, S.~Grafberger, and T.~Dunning, ``Heterogeneous data management for machine learning: A survey,'' in \emph{Proceedings of the VLDB Endowment}, vol. 14, no. 12, pp. 3324--3336, 2021.

\end{thebibliography}

