\chapter{RELATED THEORY}

\section{Machine Unlearning}
Machine unlearning refers to a class of methods that remove the influence of specific training records from a trained model while preserving utility on non-removed data. The motivation is both regulatory and operational: privacy regulations such as the General Data Protection Regulation (GDPR) and the California Consumer Privacy Act (CCPA) establish the \textit{Right to be Forgotten}, giving users the legal right to request deletion of their personal data. Beyond compliance, practical machine learning pipelines frequently require removal of mislabeled, outdated, or sensitive samples after a model has already been deployed.

In conventional machine learning, deletion is typically handled by retraining the model from scratch on the remaining dataset. While this approach is conceptually simple and guarantees complete removal of the target data's influence, it becomes impractical as models grow in size and datasets expand. Retraining a large model could take hours or days, making it impossible to honor deletion requests within reasonable timeframes, especially when multiple deletion requests arrive frequently.

Beyond computational efficiency, a practical unlearning system must address two additional concerns. First, \textit{verifiability} asks whether we can be confident that the target record no longer influences the model's predictions. Second, \textit{auditability} requires the system to maintain a clear, tamper-resistant record of deletion requests and their outcomes. In real-world deployments, this demands persistent bookkeeping structures that track active sample counts, shard assignments, and unlearning history, ensuring that model state remains consistent across repeated runs and system restarts.

\section{SISA Framework for Efficient Unlearning}
The Sharded, Isolated, Sliced, and Aggregated (SISA) framework, introduced by Bourtoule et al., provides a structured approach to efficient machine unlearning by decomposing training into smaller, independently retrainable components. The core idea is straightforward: instead of training one monolithic model on the entire dataset, the dataset is divided into multiple independent \textit{shards}. Each shard is trained separately, producing its own sub-model. When a prediction is needed, the outputs from all shard models are aggregated, typically through majority voting for classification or averaging for regression.

The key insight for unlearning is that any training record belongs to exactly one shard. Therefore, when a deletion request arrives, only the shard containing that record needs to be retrained, leaving all other shards untouched. This localized retraining dramatically reduces computational overhead compared to full model retraining. The framework can be extended further by dividing each shard into time-ordered \textit{slices}, enabling incremental learning where only the affected slice within a shard requires retraining.

From an engineering perspective, implementing SISA in a production system introduces practical requirements beyond the core algorithm. Persistent metadata storage is essential to maintain stable mappings from data indices to shards across system restarts. An append-only unlearning history log provides auditability by recording each deletion action with timestamps, identifiers, and reason codes. Idempotent unlearning behavior ensures that submitting the same deletion request multiple times does not corrupt model state or cause redundant retraining. These system-level considerations form the foundation of our implementation approach.

\section{Model Architectures Supporting Machine Unlearning}
The applicability of machine unlearning techniques varies across different model families, and the cost of unlearning depends heavily on how a model's training procedure can accommodate targeted retraining. For classical machine learning on structured data, linear models and tree-based ensembles are particularly well-suited to SISA-style sharding. These models are widely used in domains such as credit risk assessment, fraud detection, and income prediction, where both interpretability and regulatory compliance are important. The training process for these models is relatively stable, and retraining a shard can be performed independently without complex dependency management.

For neural network architectures, the landscape is more varied. Multi-Layer Perceptrons (MLPs) remain a common choice for tabular data and can be integrated into sharded training workflows with careful design. Convolutional neural networks (CNNs), such as the ResNet family, are prevalent in computer vision tasks but present additional challenges. Their training involves complex optimization dynamics and dependencies across layers, making localized retraining less straightforward to implement without careful consideration of gradient propagation and initialization strategies.

Regardless of the specific architecture, the core requirement remains consistent: unlearning must enable removal of a record's influence by retraining only the minimally affected components rather than the entire model. This project focuses primarily on scikit-learn compatible estimators for classical machine learning tasks, establishing a solid foundation that can be extended to neural architectures in future work. The theoretical principles discussed in this chapter directly inform the system design decisions, particularly around data partitioning, persistence, and audit logging.
